---
title: "Dispatch Geospatial"
output: html_notebook
---

```{r}
library(tidyverse)
library(sf) #For st_as_sf function
library(maps) #For plotting
```

```{r}
dispatch <- read_csv("data/dispatch.csv")
```

```{r}

#We need a dataframe of type sf in order to deal with geospatial data, so we'll need to convert dispatch.

#We can't have any missing values in the geo columns, so we'll have to remove them

dispatch_geo <- dispatch %>%
  drop_na(geox) %>%
  drop_na(geoy)

dispatch_geo <- st_as_sf(x = dispatch_geo,                         
           coords = c("geox", "geoy"))

#Now we have a geometry column in our dispatch_geo dataframe

```

Now that we have a geometry column, we need to declare what coordinate reference system (CRS) we're using before we can convert it into a CRS that is more useful.

The documentation is unclear about what CRS we're using. However, Boone County data shows that they are using a projection in feet and that they are using an ESRI projection (<https://maps.showmeboone.com/viewers/sd_incidentviewer/>).

A Google search of "missouri ESRI crs feet" returns ESRI:102697 as a good candidate: <https://epsg.io/102697>

We can try setting our projection to ESRI:102697.

```{r}
#Set the CRS of our dispatch_geo data to ESRI:102697

st_crs(dispatch_geo) <- "ESRI:102697"

```

```{r}
#Now convert to EPSG 4269, the most common projection for the U.S. https://epsg.io/4269 used for degrees

dispatch_geo <- st_transform(dispatch_geo, crs = 4269)

```

```{r}

#Some plotting to gut check: 

state_map_data <- map('state', fill = TRUE, plot = FALSE) %>% st_as_sf()

missouri <- state_map_data %>% 
  filter(ID == 'missouri')

ggplot() +
  geom_sf(data = missouri) +
  geom_sf(data = dispatch_geo) +
  theme_void()

```

### Converting to Zip Code

```{r}

#install.packages("ggmap")
library(ggmap)
library(devtools)
install_github('mhudecheck/revgeo')
library(revgeo)

```

First, we need a shp file of all of the U.S. ZIP codes (downloaded from here: https://catalog.data.gov/dataset/tiger-line-shapefile-2019-2010-nation-u-s-2010-census-5-digit-zip-code-tabulation-area-zcta5-na). 

This will give us polygon data for all of the ZIP codes. We're then going to perform a join to see where our points in our dispatch data intersect the polygons.

```{r}

#Read in a shp file of US zipcodes 
us_zipcodes <- read_sf('data/tl_2019_us_zcta510.shp')

#Make sure the zipcodes are in the same CRS as the dispatch data
us_zipcodes <- st_transform(us_zipcodes, 4269)

#Use smaller subset of dispatch_geo data so as not to take too much time/computing power

dispatch_geo_smaller <- dispatch_geo %>%
  slice(1:100) 

zip_codes <- dispatch_geo_smaller %>%
  st_join(us_zipcodes)

```

We could expand this to the whole dataset, but it would take a very long time to complete the join. If we were in more of a hurry, we could just subset the us_zipcodes to only Missouri zip codes, and that would speed up the join.

